---
title: "Data Preparation"
author: "Guangxin Chen"
date: "2026-04-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# Data Preparation for INFUSE

> **Note:** This guideline details the data preprocessing steps required to standardize GWAS summary statistics and reference LD panels for INFUSE

## Overview of the Pipeline

We provide a streamlined pipeline to prepare input data for INFUSE, using **All of Us (AoU)** GWAS summary statistics from European Americans and African Americans as examples. The process involves several key stages:

1. **GWAS Harmonization:** QC and filtering.
2. **LD Reference Construction:** Extracting matched genotype data from 1000 Genomes (1KG, GRCh38).
3. **LD Mismatch Detection:** Iterative pruning of variants with inconsistent LD patterns using `kriging_rss`.

All helper functions used below are defined in `code/data_prep_utils.R` and loaded in the setup chunk.

```{r setup, message=FALSE, warning=FALSE}
source(here::here("code/data_prep_utils.R"))
```

---

### 1. GWAS Harmonization

**Goal:** Standardize summary statistics from different ancestries (e.g., EUR and AFR) to a common reference panel (e.g. 1000 Genome).

We recommend removing **strand-ambiguous SNPs**, **multi-allelic variants**, **variants within the MHC region**, and variants with **minor allele frequency (MAF) < 0.01** via `GWAS_QC`. Importantly, similar to the joint-modeling approach in MESuSiE, we **restrict to the intersection of SNPs between EUR and AFR**.

```{r step1, eval=FALSE}
library(tidyverse)
source("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/utility.R")

ref   <- fread("/scratch/negishi/chen4422/UKBB_pheSVD-main/ukb_geno/in_sample_ld/1kg_build38/flipped_finemap_allbyall/1kg_afr_maf_unrelated_biallelic_nodup_extracted.bim")
colnames(ref) <- c("CHR", "SNP", "CM", "POS", "ALT", "REF")
ref   <- GWAS_QC(ref, 0.01)
annot <- fread("/scratch/negishi/chen4422/UKBB_pheSVD-main/ukb_geno/in_sample_ld/baselineLF_v2.2.UKB/annot_combine/first_three_columns.annot")

afr_file <- "/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/DBP_v8_GWAS_afr_regenie_results_From5_20250417_combined_format_noMHC.csv.gz"
eur_file <- "/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/DBP_v8_GWAS_eur_regenie_results_From5_20250417_combined_format_noMHC.csv.gz"

# QC and filter each ancestry against the 1KG reference
afrdata_qced <- read_sumstats(afr_file, ref)
eurdata_qced <- read_sumstats(eur_file, ref)

# Restrict to common SNPs, align alleles, and filter to annotation SNPs
common_snp        <- find_common_snps(eurdata_qced, afrdata_qced)
eurdata_tofinemap <- eurdata_qced %>% filter(CHR_POS %in% common_snp)
eurdata_tofinemap$SNP <- ref$SNP[match(eurdata_tofinemap$CHR_POS, ref$CHR_POS)]
afrdata_tofinemap <- allele_flip(eurdata_tofinemap,
                                 afrdata_qced %>% filter(CHR_POS %in% common_snp))
afrdata_tofinemap$SNP <- ref$SNP[match(afrdata_tofinemap$CHR_POS, ref$CHR_POS)]
eurdata_tofinemap <- eurdata_tofinemap %>% filter(SNP %in% annot$SNP)
afrdata_tofinemap <- afrdata_tofinemap %>% filter(SNP %in% annot$SNP)

# Write outputs
trait   <- "DBP"
out_dir <- "/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/summstats_to_finemap/"
fwrite(eurdata_tofinemap,paste0(out_dir, trait, "_EUR_GRCh38_combined_format_noMHC_to_finemap.csv"))
fwrite(afrdata_tofinemap,paste0(out_dir, trait, "_AFR_GRCh38_combined_format_noMHC_to_finemap.csv"))
fwrite(eurdata_tofinemap[, c(1, 6)], paste0(out_dir, trait, "_GRCh38_combined_format_noMHC_to_finemap_ref_flip"),  col.names = FALSE, sep = " ")
fwrite(eurdata_tofinemap[, c(1)], paste0(out_dir, trait, "_GRCh38_combined_format_noMHC_to_finemap_snplist"),   col.names = FALSE)
```

Note that the **ALT and REF alleles need to be consistent across ancestry groups**. In the *All of Us* dataset this is pre-harmonized; for other sources see [this document](https://borangao.github.io/meSuSie_Analysis/GWAS_QC.html).

---

### 2. Align GWAS Summary Statistics with LD Reference Panel

**Goal:** Construct ancestry-matched LD reference genotypes from 1KG that **exactly match** the GWAS variant list and use the same REF/ALT orientation, avoiding sign errors in Z-scores.

We use `plink2` to subset 1KG genotypes (`--extract`), enforce REF allele orientation (`--ref-allele`), and output a new `.bed/.bim/.fam` file (`--make-bed`).

```{r step2, eval=FALSE}
trait <- "DBP"
align_ld_ref("EUR", trait)
align_ld_ref("AFR", trait)
```

---

### 3. LD Reference Construction

Below, we use the region spanning **58,658,279–59,673,534 bp on chromosome 20** (DBP) as an illustrative example. `build_ld_matrix()` extracts SNPs in the region, runs plink2, flags monomorphic variants, and writes the LD correlation matrix.

```{r step3, eval=FALSE}
trait   <- "DBP"
kg_dir  <- paste0("/scratch/negishi/chen4422/UKBB_pheSVD-main/ukb_geno/in_sample_ld/1kg_build38/flipped_finemap_allbyall/lipids_v8/test_MF/")
eur_ref <- fread(paste0(kg_dir, "1kg_eur_maf_unrelated_biallelic_nodup_extracted_", trait, "_flipped.bim"))
afr_ref <- fread(paste0(kg_dir, "1kg_afr_maf_unrelated_biallelic_nodup_extracted_", trait, "_flipped.bim"))

start <- 58658279; end <- 59673534

build_ld_matrix(eur_ref, chr_num = 20, start_pos = start, end_pos = end,
                bfile   = paste0(kg_dir, "1kg_eur_maf_unrelated_biallelic_nodup_extracted_", trait, "_flipped"),
                out_dir = paste0("risk_loci_ld_eur_", trait))

build_ld_matrix(afr_ref, chr_num = 20, start_pos = start, end_pos = end,
                bfile   = paste0(kg_dir, "1kg_afr_maf_unrelated_biallelic_nodup_extracted_", trait, "_flipped"),
                out_dir = paste0("risk_loci_ld_afr_", trait))
```

---

### 4. LD Mismatch Detection

A common issue in fine-mapping is mismatch between summary statistics and the LD reference panel. We use `kriging_rss` from `susieR` to iteratively identify and remove SNPs whose Z-scores are inconsistent with the LD structure. `filter_ld_mismatch()` handles the full loop automatically.

```{r step4, message=FALSE, warning=FALSE}
trait <- "DBP"

geno_dir_eur <- paste0("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/risk_loci_ld_eur_", trait, "/")
geno_dir_afr <- paste0("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/risk_loci_ld_afr_", trait, "/")

afr_sumstats <- fread(paste0("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/summstats_to_finemap/", trait, "_AFR_GRCh38_combined_format_noMHC_to_finemap.csv"))
eur_sumstats <- fread(paste0("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/summstats_to_finemap/", trait, "_EUR_GRCh38_combined_format_noMHC_to_finemap.csv"))

afr_res <- filter_ld_mismatch(geno_dir_afr, "loci_example", afr_sumstats)
```

The **Kriging RSS diagnostic** plot color-codes SNPs by their likelihood of being an LD outlier. In the example above, an initial outlier was identified and removed in the following iteration.

We apply the same procedure to the European American group, then save the cleaned outputs:

```{r step4-eur, eval=FALSE}
eur_res <- filter_ld_mismatch(geno_dir_eur, "loci_example", eur_sumstats)

data_dir <- paste0("/scratch/negishi/chen4422/hapnest/multi-ans-sum-stats/Lipids_all_of_us_v8/formatted/format_summary_stats/results_", trait, "/summary_data/")
fwrite(afr_res$sumstats, paste0(data_dir, "afr_loci_example_summ"))
fwrite(eur_res$sumstats, paste0(data_dir, "eur_loci_example_summ"))
fwrite(afr_res$cov,      paste0(data_dir, "afr_loci_example.ld"), sep = " ")
fwrite(eur_res$cov,      paste0(data_dir, "eur_loci_example.ld"), sep = " ")
```
